Outpacing the Lanes: Toronto's Bike Share Ridership Surges Ahead of Infrastructure Growth

### 2024-09-26 paper/paper.qmd
Final edits to improve written portions of paper

This study utilizes two primary datasets: Toronto's Bike Share ridership data 
[@bikeShareData] and the city's Cycling Network (Bikeways) data 
[@bikeWayData]. Both datasets were obtained from the City of Toronto's Open Data Portal [@torontoOpenData], providing a comprehensive view of cycling infrastructure from 2001 to 2024, and Bike Share ridership in the city from 2017 to 2023.

@@

We cleaned the data by downloading all CSV files, extracting required columns (trip ID and start date), and performing necessary cleaning operations. This process involved standardizing column names across files, converting data types, and aggregating the data to calculate monthly total ridership figures. The cleaned data provides a comprehensive view of Bike Share usage from January 2017 to December 2023.

One challenge in processing this data was the inconsistency in column names and data formats across different years. This was addressed through a custom R function that standardized column names and data types across all files. Additionally, any incomplete or invalid entries were removed during the cleaning process to ensure data integrity.

A sample of this processed dataset can be seen in (@tbl-sample-bikeshare), showcasing the year and month of observation along with the total number of rides for that month. This granular monthly data allows for analysis of seasonal trends and year-over-year growth in ridership.

@@

To provide an overview of the Bike Share program's growth, we aggregated the monthly data into annual statistics, as shown in (@tbl-ridership_summary_stats).

@@

This summary reveals a consistent increase in total annual rides from 1,190,320 in 2017 to 5,713,141 in 2023, representing nearly a fivefold increase over seven years. The average monthly rides have grown correspondingly, indicating a substantial expansion of the program's usage.

## Cycling Network (Bikeways) Data
The Cycling Network data was sourced from the City of Toronto's Transportation Services division. This dataset provides detailed information about the city's bikeways, including their construction year, any upgrades, and the type of bikeway. (@tbl-sample_bikeway) offers a sample of this dataset.

@@

1. **Protected Lanes**: Physically separated and protected from motor vehicle traffic. This includes Bi-Direction Cycle Tracks, Cycle Tracks, and Multi-Use Trails.
2. **On-Road Lanes**: Designated lanes on the road, but not physically separated. This category encompasses various types of bike lanes.
3. **Shared Roadways**: Roads where cyclists and motor vehicles share the same space. This includes Sharrows, Signed Routes, and Park Roads.

It's important to note that in the original dataset, bikeways constructed before 2001 were all labeled with a construction year of 2001. This data limitation could potentially lead to skewing or misunderstanding in the analysis, particularly for earlier years. Additionally, some entries with invalid years (e.g., year labeled as "1") or missing data were filtered out during the cleaning process to ensure data quality.

Another key limitation to note about the bikeway data, is that our analysis relies on the count of bikeways rather than their total length. This approach was necessitated by data availability constraints, as information on the length of each bikeway was not provided in the dataset. Using counts instead of lengths may not provide a fully accurate representation of infrastructure growth, as it doesn't account for variations in bikeway size or capacity. For instance, a short protected lane would be given the same weight in our analysis as a long multi-use trail. This limitation should be kept in mind when interpreting the results, particularly when comparing infrastructure growth to ridership increases.

Both datasets were cleaned and processed using R [@RCoreTeam2023], with additional support from the tidyverse [@tidyverse], dplyr [@dplyr], and lubridate [@lubridate] packages. The cleaning process involved removing any incomplete entries, standardizing date formats for consistency across the datasets, and aggregating data where necessary.

These datasets, chosen for their up-to-date and comprehensive nature, offer the most accurate available information on Toronto's cycling infrastructure and Bike Share usage. When analyzed together, they provide insights into the relationship between infrastructure development and the growing demand for cycling in Toronto, allowing us to assess whether the city's bikeway expansion is keeping pace with increasing ridership. While other datasets were considered, such as bike lanes measured in kilometers or total bike counts, they were not used due to being outdated or no longer maintained.

@@

## Bike Share Ridership Trends
Toronto's Bike Share monthly ridership from 2017 to 2023 is illustrated in (@fig-ridership_monthly_graph), showing a clear upward trajectory with pronounced seasonal fluctuations. Peak usage consistently occurs during the summer months, while winter sees a significant reduction in rides. Despite these seasonal variations, the overall trend line demonstrates steady growth in ridership.

@@

## Cycling Infrastructure Development
A breakdown of bikeway types constructed each year from 2002 to 2023 is 
presented by (@fig-bikeway_type_constructed_graph). This graph reveals several 
interesting patterns in the development of Toronto's cycling infrastructure:

1. A significant spike in bikeway installations occurred between 2005 and 2007, primarily driven by the addition of shared roadways. This surge aligns with the city's efforts to implement its 2001 Bike Plan, which aimed to create 1,000 km of bikeways by 2011. While the city fell short of this ambitious target, the period saw a substantial increase in shared roadways, adding over 145 km by 2007 [@ratchford]. As shown in (@fig-cumulative-proportion-bikeways), this rapid expansion of shared roadways significantly altered the proportional makeup of Toronto's bikeway network, with shared roadways dominating the infrastructure mix for several years following 2007.
2. After 2007, there's a noticeable shift towards more protected lanes and on-road lanes, reflecting growing concerns about cyclist safety and calls for more separated cycling infrastructure.
3. The installation of new bikeways has been relatively consistent from 2010 
onwards, with a mix of all three types (shared roadways, on-road lanes, and 
protected lanes) being added each year.

@@

Bikeways were also upgraded throughout the years to improve the quality and
safety for bikers.(@fig-bikeway_type_upgraded_graph) shows the breakdown of 
bikeway upgrades by year from 2002 to 2023. Notable observations include:

@@

## Cumulative Growth of Cycling Infrastructure
To provide a complete understanding of Toronto's entire Cycling network, 
(@fig-cumulative-growth) presents the cumulative growth of different bikeway 
types from 2001 to 2023, demonstrating:

@@

\newpage
## Comparison of Ridership and Infrastructure Growth
To directly compare the ridership and infrastructure growth, (@fig-combined_ridership_bikeways_graph) provides a comparative view of annual 
Bike Share ridership and the cumulative number of bikeways from 2017 to 2023.
This graph reveals:

@@

ggplot(combined_data, aes(x = year)) +
  geom_bar(aes(y = cumulative_bikeways * 2000, fill = "Bikeways"), 
           stat = "identity", alpha = 0.75) +
  geom_line(aes(y = total_rides, color = "Ridership"), size = 1) +
  scale_y_continuous(
    name = "Total Annual Rides", 
    sec.axis = sec_axis(~ . / 2000, name = "Cumulative Number of Bikeways"),
    labels = function(x) paste0(x/1000000, "M"),
    breaks = seq(0, max(combined_data$total_rides), by = 1000000),
    minor_breaks = seq(0, max(combined_data$total_rides), by = 500000)
  ) +
  scale_x_continuous(
    breaks = seq(2017, 2023, by = 1),
    limits = c(2016.5, 2023.5),
    expand = c(0, 0)
  ) +
  scale_color_manual(values = c("Ridership" = "#1f77b4")) +
  scale_fill_manual(values = c("Bikeways" = "gray50")) +
  labs(
    x = "Year",
    color = NULL,
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.major = element_line(color = "gray85", linewidth=0.35)
  )
```

>>>>

This study utilizes two primary datasets: Toronto's Bike Share ridership data [@bikeShareData] and the city's Cycling Network (Bikeways) data [@bikeWayData]. Both datasets were obtained from the City of Toronto's Open Data Portal [@torontoOpenData], providing a comprehensive view of cycling infrastructure from 2001 to 2024, and Bike Share ridership in the city from 2017 to 2023. These datasets were selected for their up-to-date and comprehensive nature, offering the most accurate available information on Toronto's cycling infrastructure and Bike Share usage.

@@

We cleaned the data by downloading all CSV files, extracting required columns (trip ID and start date), and performing necessary cleaning operations. This process involved standardizing column names across files, standardizing data types, and aggregating the data into monthly total ridership figures. Additionally, any incomplete or invalid entries were removed ruing the cleaning process to ensure data integrity. This process resulted in a cleaned dataset providing a comprehensive view of Bike Share usage from January 2017 to December 2023.

@tbl-sample-bikeshare presents a sample of the processed dataset, showcasing the year and month of observation along with the total number of rides for that month. This granular monthly data allows for analysis of seasonal trends and year-over-year growth in ridership.

@@

To provide an overview of the Bike Share program's growth, we aggregated the monthly data into annual statistics, as shown in @tbl-ridership_summary_stats.

@@

This summary reveals a consistent increase in total annual rides from 1,190,320 in 2017 to 5,713,141 in 2023, representing nearly a fivefold increase over seven years. The average monthly rides have grown correspondingly, indicating a substantial expansion of the bike shares program's usage.

## Cycling Network (Bikeways) Data
The Cycling Network data, sourced from the City of Toronto's Transportation Services division, provides detailed information about the city's bikeways. This includes their construction year, any upgrades, and the type of bikeway. @tbl-sample_bikeway offers a sample of this dataset.

@@

1. **Protected Lanes**: Physically separated and protected from motor vehicle traffic (e.g., Bi-Direction Cycle Tracks, Cycle Tracks, Multi-Use Trails).
2. **On-Road Lanes**: Designated lanes on the road, not physically separated (various types of bike lanes).
3. **Shared Roadways**: Roads where cyclists and motor vehicles share the same space (e.g., Sharrows, Signed Routes, Park Roads).

The data cleaning process for the Cycling Network dataset involved:

1. Filtering out Constructed year and Upgraded year entries with invalid years (e.g., year labeled as "1") or missing years.
2. Standardizing bikeway type classifications into the three main categories (Protected Lanes, On-Road Lanes, Shared Roadways).
3. Addressing the limitation that bikeways constructed before 2001 were all labeled with a construction year of 2001 in the original dataset.

## Measurement and Limitations
It's crucial to note that our analysis relies on the count of bikeways rather than their total length. This approach was necessitated by data availability constraints, as information on the length of each bikeway was not provided in the dataset. Using counts instead of lengths may not provide a fully accurate representation of infrastructure growth, as it doesn't account for variations in bikeway size or capacity. For instance, a short protected lane would be given the same weight in our analysis as a long multi-use trail.

This limitation should be kept in mind when interpreting the results, particularly when comparing infrastructure growth to ridership increases. Future studies with access to length data could offer a more precise comparison between infrastructure development and ridership growth.

## Data Processing and Analysis
Both datasets were cleaned and processed using R [@RCoreTeam2023], with additional support from the tidyverse [@tidyverse], dplyr [@dplyr], and lubridate [@lubridate] packages. The cleaning process involved removing any incomplete entries, standardizing date formats for consistency across the datasets, and aggregating data where necessary.

When analyzed together, these datasets provide insights into the relationship between infrastructure development and the growing demand for cycling in Toronto, allowing us to assess whether the city's bikeway expansion is keeping pace with increasing ridership. While other datasets were considered, such as bike lanes measured in kilometers or total bike counts, they were not used due to being outdated or no longer maintained.

@@

## Bike Share Ridership Trends
Toronto's Bike Share monthly ridership from 2017 to 2023 is illustrated in @fig-ridership_monthly_graph, showing a clear upward trajectory with pronounced seasonal fluctuations. Peak usage consistently occurs during the summer months, while winter sees a significant reduction in rides. Despite these seasonal variations, the overall trend line demonstrates steady growth in ridership.

@@

## Cycling Infrastructure Development
A breakdown of bikeway types constructed each year from 2002 to 2023 is 
presented by @fig-bikeway_type_constructed_graph. This graph reveals several 
interesting patterns in the development of Toronto's cycling infrastructure:

1. A significant spike in bikeway installations occurred between 2005 and 2007, primarily driven by the addition of shared roadways. This surge aligns with the city's efforts to implement its 2001 Bike Plan, which aimed to create 1,000 km of bikeways by 2011. While the city fell short of this ambitious target, the period saw a substantial increase in shared roadways, adding over 145 km by 2007 [@ratchford]. As shown in @fig-cumulative-proportion-bikeways, this rapid expansion of shared roadways significantly altered the proportional makeup of Toronto's bikeway network, with shared roadways dominating the infrastructure mix for several years following 2007.
2. After 2007, there's a noticeable shift towards more protected lanes and on-road lanes, reflecting growing concerns about cyclist safety and calls for more separated cycling infrastructure.
3. The installation of new bikeways has been relatively consistent from 2010 
onwards, with a mix of all three types (shared roadways, on-road lanes, and 
protected lanes) being added each year.

@@

Bikeways were also upgraded throughout the years to improve the quality and
safety for bikers. @fig-bikeway_type_upgraded_graph shows the breakdown of 
bikeway upgrades by year from 2002 to 2023. Notable observations include:

@@

## Cumulative Growth of Cycling Infrastructure
To provide a complete understanding of Toronto's entire Cycling network, 
@fig-cumulative-growth presents the cumulative growth of different bikeway 
types from 2001 to 2023, demonstrating:

@@

## Comparison of Ridership and Infrastructure Growth
To directly compare the ridership and infrastructure growth, @fig-combined_ridership_bikeways_graph provides a comparative view of annual 
Bike Share ridership and the cumulative number of bikeways from 2017 to 2023.
This graph reveals:

@@

ggplot(combined_data, aes(x = year)) +
  geom_bar(aes(y = cumulative_bikeways * 2000, fill = "Bikeways"), 
           stat = "identity", alpha = 0.75) +
  geom_line(aes(y = total_rides, color = "Ridership"), size = 1) +
  scale_y_continuous(
    name = "Total Annual Rides", 
    sec.axis = sec_axis(~ . / 2000, name = "Cumulative Number of Bikeways"),
    labels = function(x) paste0(x/1000000, "M"),
    breaks = seq(0, max(combined_data$total_rides), by = 1000000),
    minor_breaks = seq(0, max(combined_data$total_rides), by = 500000)
  ) +
  scale_x_continuous(
    breaks = seq(2017, 2023, by = 1),
    limits = c(2016.5, 2023.5),
    expand = c(0, 0)
  ) +
  scale_color_manual(values = c("Ridership" = "#1f77b4")) +
  scale_fill_manual(values = c("Bikeways" = "gray40")) +
  labs(
    x = "Year",
    color = NULL,
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.major = element_line(color = "gray85", linewidth=0.35)
  )
```
